Parallel Multi-objective optimization using co-evolutionary islands and search space overlapping

### 2016-06-14 hpmoon-hpca.tex
Minor changes

Different approaches have been used to parallelize Evolutionary Algorithms (EAs), as each individual can be considered as an independient unit \citep{Alba13parallel}. Classic methods, such as the global parallel EAs (Master-Slave), or the spatially structured algorithms (Island model or Cellular EAs) have been applied successfully in the past \citep{Folino03cellular,Alba02Parallelism}.  In the case of Multi-Objective Evolutionary Algorithms (MOEAs), these approaches \citep{Luna15Survey} need to deal with a whole solution set called Pareto Front (PFs). This imply to use different distribution and sharing mechanisms, as there exist a tension between the speedup achievable from parallelization and the need to globally recombine results to accurately identify the PF \citep{Branke04Parallelizingcone}.

The application of MOEAs to solve real-world problems have gained attention in the last years \citep{Luna15Survey,Mukhopadhyay14Survey,Chavez15MO,Hidalgo16residualstress}. These kind of problems require a high number of variables, meaning that the EAs need to deal with large individuals and spent significant extra time for crossover, mutation and migration. Different authors have proposed methods to divide the decision space (the chromosome) to improve the performance and solution quality. In this aspect, the co-evolution model is a dimension-distributed model where a high-dimensional problem is divided into lower dimensional problems \citep{Gong15models,Tonda12cooperative}, and evolved separately. One example of application of this technique was described in \citep{Kimovski15Parallel}. The method presented in that works involves different workers that evolve sub-populations created and recombined by a master process, which performs different recombination alternatives from the parts returned by worker processes. A high dimensional problem was used to compare these
alternatives. In \citep{Dorronsoro13superlinear}, a distributed coevolutionary island model was used.  Although in both previous works significant speedups were attained, only a low number of nodes were used (8). However, when increasing the number of islands, the division of each section of the chromosome becomes smaller, and the scalability may be affected by obtaining lower quality solutions in the same amount of time.

@@

Evolutionary algorithms (EAs) \citep{DBLP:series/ncs/EibenS15,DeJong2006} are bio-inspired meta-heuristics that can be effectively used to find nearly optimal solutions for optimization problems. Usually an EA starts by generating a set of random solutions, called \emph{population}, following a user-defined description. Then, it evaluates each candidate solution, called \emph{individual}, assigning it a \emph{fitness} value, that describes how good the individual is, with regards to the target problem. New solutions are then generated by the application of \emph{operators} that either mutate a single existing solutions or recombine different existing solutions. After each iteration, called \emph{generation}, the least fit individuals are removed, and the process continues until a user-defined stop condition is met.

Multi-objective optimization problems (MOP) are those where several objectives have to be simultaneously optimized \cite{Mora13paretobased}. Solving a MOP implies maximize or minimize a function composed by several cost functions (one per objective). In these problems the aim is to obtain a set of solutions that are better than the rest considering all the objectives; this set is known as the Pareto Front. The solutions in this set are {\em non-dominated}, that is, none is better than the others for all the objectives. MOEAs are sub-set of EAs used to solve MOPs. One of the most well-known and referenced MOEAs \cite{Dorronsoro13superlinear} is the {\em Non-dominated Sorting Genetic Algorithm} (NSGA-II) \cite{Deb00NSGAII}. Classical genetic operators are applied to all the individuals, and divided in different ranks (based on the dominance) to be selected for next generations.

Since the early 2000s, the distributed and parallel Evolutionary Algorithms have been used, mostly to take the most from systems such as clusters or grids \citep{Talbi08Parallel}. But in the case of MOEAs, the distribution and parallelization is not as direct as in single-objective EAs. This is because in different steps of the algorithm a whole set of dependent solutions, the Pareto Front, should be managed as a whole, spending time in gather all individuals from the different process or islands.

To solve this issue, some authors have proposed the usage of Master-Slave approaches. For example,  \cite{Durillo08masterslave} compared different master-slave approaches {\color{red} TODO DESCRIBIR MEJOR ESTE TRABAJO}. The method proposed by Hiroyasu \citep{Hiroyasu07discussion} generates offspring depending on the computation power {\color{red} TODO Y ESTE}.

The work by \cite{Deb03distributed} was one of the first approaches for distributed MOEAs (dMOEAs). In that work, the dominance of solutions was divided in the islands using a transformation of coordinates. Their authors concluded that dividing the search space is a good idea, but achieve this is not trivial. Other approaches for dividing the search space include objective separation in different processors \citep{Xiao03specialized}, or divide in two populations: one for elite and other search \citep{Wang09parallel}. Martens, in \citep{Martens13asynchronous} generated a Barabasi-Albert network as the island topology, and selection based on migrating individuals from a not-crowded area and acceptance based on diversity.

@@

In our previous work \citep{Garcia16hpmoon} we used some of the previous ideas to compare two co-evolutionary dMOEAs. The first one divided the chromosome in $P$ sections, being $P$ the number of islands. Every island $p$ only performed the mutation and crossover in that part ($p_{th}$) of the chromosome, while the fitness was calculated using the whole individual. After certain number of generations, individuals were migrated randomly to other islands. The performance metrics were calculated at the end of the run. The second method, coevolution with overlapping islands, used the $p_{th-1}$ and $p_{th+1}$ sections of each chromosome, besides the $p_{th}$ part. During the same amount of time both methods obtained better results than a baseline algorithm that dealt with the whole chromosome in each island for crossover and mutation. We discovered that the performance using one or another method depends on the number of sections of the individuals and the number of islands used. This motivates us to find a new adaptive method to select this number of sections of the chromosome to use, depending on the number of islands. Moreover, previous experiments were performed in a single processor synchronous island model with a limited number of islands (8, 32 and 128). In this paper, we have used 8, 16, 32, 64 and 128 islands, and this time, using a real cluster. Therefore, at the same time we are proposing an adaptive method to divide the individual search space, we are validating the previous methods.

@@

\begin{figure*}
\centering
\includegraphics[width=12cm]{islandNoDisjoint.jpg}
\caption{{\color{red} NOTA: ESTA FIGURA TIENE QUE ACTUALIZARSE CON LA ADAPTATIVA!!!!!} Adaptive overlapping algorithm: every island $p$ modifies the  $p+c$,
  $p_{th}$ and $p-c$  components (in grey) of the individuals. $c$ is calculated depending on the number of islands.}
  \label{fig:adaptive}
\end{figure*}
% --------------------------------------------------------------

@@

Different metrics, explained in the previous section, have been used to calculate the quality of the obtained PFs in each configuration. As some of the metrics  (such as HV) require a reference point to be calculated, the point (1,9) has been chosen as reference, as none of the generated PFs in all runs are dominated by it. Metrics are then normalized with respect to that point. A Kruskall-Wallis significance test has been performed to the metrics of all runs of the configurations, as the Kolmogorov-Smirnov test detected non-normal distributions. The average results for each configuration are shown in Table \ref{tab:results512} (for 512 dimensions) and \ref{tab:results2048} (for 2048).

@@

%DESCRIPTION OF THE RESULTS
Results show that there is a difference in performance when using 512 or 2048 dimensions. Although in some cases the overlapped methods obtain quality metrics values better or significantly equal than the obtained by the baseline, it is clear that 512 dimensions is not a value high enough to apply these methods. This contradicts the results obtained in our previous work in a shared-memory mono-processor version \citep{Garcia16hpmoon}, where the overlapping method obtained best results than the baseline in both dimension lengths. This can be explained because the added migration latency between nodes requires more time than the time saved during the crossover and migration.

@@

%EXPLANATION OF THE RESULTS
This can be explained comparing the number non-dominated solutions in each island and the average number of generations (Table \ref{tab:sols512} and Table \ref{tab:sols2048}). With both chromosome lengths, increasing the number of islands implies more generations with all the methods (logically, as there are less individuals in each island). But also, the co-evolutionary methods approach to the number of generations with the baseline when increasing the number of islands. However, with $L=2048$ the migration time is enough big to almost not improve the number of generations with respect to the baseline, even in the lower number of islands, as it happens with $L=512$. Therefore, more generations do not necessarily improve the solution of the global PF, but focusing in different elements of the chromosome. As indicated previously, every island is unaware of the other islands PFs, and they are trying to optimize their solutions independently. With respect to the average number of solution in each island, there is a clear difference in number with the baseline with $L=512$, where this value is in the most of the cases, less than a half. The number of non-dominated solutions also improves a better Spread indicator, where the baseline almost always obtains better values %TENGO QUE COMENTAR MAS LOS RESULTADOS Y ENTENDERLO TODO Y JUSTIFICARLO FINAMENTE

Finally, as happened in \citep{Garcia16hpmoon}, results show that all the quality indicators decrease when the number of islands is increased, as the the sizes of the subpopulations are smaller. This is consistent with the claim by \cite{Dorronsoro13superlinear}, who found that cooperative coevolutionary MOEAs work better on bigger populations (more than 100 individuals).

>>>>

Different approaches have been used to parallelize Evolutionary Algorithms (EAs), as each individual can be considered as an independent unit \citep{Alba13parallel}. Classic methods, such as the global parallel EAs (Master-Slave), or the spatially structured algorithms (Island model or Cellular EAs) have been applied successfully in the past \citep{Folino03cellular,Alba02Parallelism}.  In the case of Multi-Objective Evolutionary Algorithms (MOEAs), these approaches \citep{Luna15Survey} need to deal with a whole solution set called Pareto Front (PFs). This imply to use different distribution and sharing mechanisms, as there exist a tension between the speedup achievable from parallelization and the need to globally recombine the results to accurately identify the PF \citep{Branke04Parallelizingcone}.

The application of MOEAs to solve real-world problems have gained attention in the last years \citep{Luna15Survey,Mukhopadhyay14Survey,Chavez15MO,Hidalgo16residualstress}. These kind of problems requires a high number of variables, meaning that the EAs need to deal with large individuals and spent significant extra time for crossover, mutation and migration. Different authors have proposed methods to divide the decision space (the chromosome) to improve the performance and solution quality. In this aspect, the co-evolution model is a dimension-distributed model where a high-dimensional problem is divided into lower dimensional problems \citep{Gong15models,Tonda12cooperative}, and evolved separately. One example of application of this technique was described in \citep{Kimovski15Parallel}. The method presented in that works involves different workers that evolve sub-populations created and recombined by a master process, which performs different recombination alternatives from the parts returned by worker processes. A high dimensional problem was used to compare these
alternatives. In \citep{Dorronsoro13superlinear}, a distributed coevolutionary island model was used.  Although in both previous works significant speedups were attained, only a low number of nodes were used (8). However, when increasing the number of islands, the division of each section of the chromosome becomes smaller, and the scalability may be affected by obtaining lower quality solutions in the same amount of time.

@@

Evolutionary algorithms (EAs) \citep{DBLP:series/ncs/EibenS15,DeJong2006} are bio-inspired meta-heuristics that can be effectively used to find nearly optimal solutions for optimization problems. Usually an EA starts by generating a set of random solutions, called \emph{population}, following a user-defined description. Then, it evaluates each candidate solution, called \emph{individual}, assigning it a \emph{fitness} value, that describes how good the individual is, with regards to the target problem. New solutions are then generated by the application of \emph{operators} that either mutate a single solution or recombine different existing solutions. After each iteration, called \emph{generation}, the least fit individuals are removed, and the process continues until a user-defined stop condition is met.

Multi-objective optimization problems (MOP) are those where several objectives have to be simultaneously optimized \cite{Mora13paretobased}. Solving a MOP implies maximize or minimize a function composed of several cost functions (one per objective). In these problems the aim is to obtain a set of solutions that are better than the rest, considering all the objectives; this set is known as the Pareto Front. The solutions in this set are {\em non-dominated}, that is, none is better than the others for all the objectives. MOEAs are sub-set of EAs used to solve MOPs. One of the most well-known and referenced MOEAs \cite{Dorronsoro13superlinear} is the {\em Non-dominated Sorting Genetic Algorithm} (NSGA-II) \cite{Deb00NSGAII}. Classical genetic operators are applied to all the individuals, and divided into different ranks (based on the dominance) to be selected for the next generations.

Since the early 2000s, the distributed and parallel Evolutionary Algorithms have been used, mostly to take the most from systems such as clusters or grids \citep{Talbi08Parallel}. But in the case of MOEAs, the distribution and parallelization is not as direct as in single-objective EAs. This is because in different steps of the algorithm a whole set of dependent solutions, the Pareto Front, should be managed as a whole, spending time in gathering all individuals from the different processors or islands.

To solve this issue, some authors have proposed the usage of Master-Slave approaches. For example,  \cite{Durillo08masterslave} compared different master-slave approaches {\color{red} TODO DESCRIBIR MEJOR ESTE TRABAJO}. The method proposed by Hiroyasu \citep{Hiroyasu07discussion} generates offspring depending on the computational power {\color{red} TODO Y ESTE}.

The work by \cite{Deb03distributed} was one of the first approaches for distributed MOEAs (dMOEAs). In that work, the dominance of solutions was divided into the islands using a transformation of coordinates. Their authors concluded that dividing the search space is a good idea, but achieve this is not trivial. Other approaches for dividing the search space include objective separation in different processors \citep{Xiao03specialized}, or divide in two populations: one for elite and other search \citep{Wang09parallel}. Martens, in \citep{Martens13asynchronous} generated a Barabasi-Albert network as the island topology, and selection based on migrating individuals from a not-crowded area and acceptance based on diversity.

@@

In our previous work \citep{Garcia16hpmoon} we used some of the previous ideas to compare two co-evolutionary dMOEAs. The first one divided the chromosome in $P$ sections, being $P$ the number of islands. Every island $p$ only performed the mutation and crossover in that part ($p_{th}$) of the chromosome, while the fitness was calculated using the whole individual. After a certain number of generations, individuals were migrated randomly to other islands. The performance metrics were calculated at the end of the run. The second method, coevolution with overlapping islands, used the $p_{th-1}$ and $p_{th+1}$ sections of each chromosome, besides the $p_{th}$ part. During the same amount of time both methods obtained better results than a baseline algorithm that dealt with the whole chromosome in each island for crossover and mutation. We discovered that the performance using one or another method depends on the number of sections of the individuals and the number of islands used. This motivates us to find a new adaptive method to select this number of sections of the chromosome to use, depending on the number of islands. Moreover, previous experiments were performed on a single processor synchronous island model with a limited number of islands (8, 32 and 128). In this paper, we have used 8, 16, 32, 64 and 128 islands, and this time, using a real cluster. Therefore, at the same time we are proposing an adaptive method to divide the individual search space, we are validating the previous methods.

@@

\begin{figure*}
\centering
\includegraphics[width=12cm]{islandAdaptive.jpg}
\caption{Adaptive overlapping algorithm: every island $p$ modifies the  $p+c$,
  $p_{th}$ and $p-c$  components (in grey) of the individuals. $c$ is calculated depending on the number of islands. In this case, $c=2$.}
  \label{fig:adaptive}
\end{figure*}
% --------------------------------------------------------------

@@

Different metrics, explained in the previous section, have been used to calculate the quality of the obtained PFs in each configuration. As some of the metrics  (such as HV) require a reference point to be calculated, the point (1,9) has been chosen as reference, as none of the generated PFs in all runs are dominated by it. Metrics are then normalized with respect to that point. A Kruskall-Wallis significance test has been performed to the metrics of all runs of the configurations, as the Kolmogorov-Smirnov test detected non-normal distributions. The average results for each configuration are shown in Table \ref{tab:results512} (for 512 dimensions) and \ref{tab:results2048} (for 2048). As previously explained, when $P=8$ the results of A are equivalent to the obtained with D (because $c=0$), and when $P=16$ they are equivalent to O ($c=1$).

@@

%DESCRIPTION OF THE RESULTS
Results show that there is a difference in performance when using 512 or 2048 dimensions. Although in some cases the overlapped methods obtain quality metrics values better or significantly equal than the obtained by the baseline, it is clear that 512 dimensions is not a value high enough to apply these methods. This contradicts the results obtained in our previous work in a shared-memory mono-processor version \citep{Garcia16hpmoon}, where the overlapping method obtained better results than the baseline in both dimension lengths. This can be explained because the added migration latency between nodes requires more time than the time saved during the crossover and migration.

@@

%EXPLANATION OF THE RESULTS
This can be explained comparing the number non-dominated solutions in each island and the average number of generations (Table \ref{tab:sols512} and Table \ref{tab:sols2048}). With both chromosome lengths, increasing the number of islands implies more generations with all the methods (logically, as there are less individuals in each island). But also, the co-evolutionary methods approach to the number of generations with the baseline when increasing the number of islands. However, with $L=2048$ the migration time is enough big to almost not improve the number of generations with respect to the baseline, even in the lower number of islands, as it happens with $L=512$. Therefore, more generations do not necessarily improve the solution of the global PF, but focusing on different elements of the chromosome. As indicated previously, every island is unaware of the other islands PFs, and they are trying to optimize their solutions independently. With respect to the average number of solutions in each island, there is a clear difference in number with the baseline with $L=512$, where this value is in the most of the cases, less than a half. The number of non-dominated solutions also improves a better Spread indicator, where the baseline almost always obtains better values %TENGO QUE COMENTAR MAS LOS RESULTADOS Y ENTENDERLO TODO Y JUSTIFICARLO FINAMENTE

Finally, as happened in \citep{Garcia16hpmoon}, results show that all the quality indicators decrease when the number of islands is increased, as the  sizes of the subpopulations are smaller. This is consistent with the claim by \cite{Dorronsoro13superlinear}, who found that cooperative coevolutionary MOEAs work better on bigger populations (more than 100 individuals).
